---
title: "STA303 Proejct Part 1: Exploratory Data Analysis"
output: html_notebook
---


```{r}
setwd("C:\\Users\\USER\\OneDrive - University of Toronto\\UofT Classes\\STA303\\STA303 Project")

music_data <- read.csv("data_by_genres.csv")

library(ggplot2)
library(lmtest)
library(car)
library(knitr) ## To create nice tables
knitr::opts_chunk$set(echo = TRUE)

## Previewing the data
head(music_data)

```


```{r}
## Getting a general overview of the data
summary(music_data)
```


```{r}
## Checking to see if dataset has any misisng values
colSums(is.na(music_data))
```

```{r}
## Confirming that each column type is correct
str(music_data)
music_data$popularity <- round(music_data$popularity) # since we want to use a Poisson distribution to model this
```
# Histograms of the distributions of the Continuous variables 

```{r}
# HISTOGRAM FOR MY TARGETED RESPONSE VARIABLE

# Histogram for Popularity Scores 
ggplot(data = music_data, aes(x = popularity)) +
  geom_histogram(binwidth = 5, fill = "purple", color = "black") + # Adjust binwidth as necessary
  ggtitle("Histogram of Popularity Scores") +
  xlab("Popularity Score") +
  ylab("Frequency")
```


```{r}
# HISTOGRAMS FOR THE CONTINUOUS PREDICTOR VARIABLES IN THIS DATASET

# Histogram for Acousticness
ggplot(data = music_data, aes(x = acousticness)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  ggtitle("Histogram of Acousticness") +
  xlab("Acousticness") +
  ylab("Frequency")

# Histogram for Duration in milliseconds
ggplot(data = music_data, aes(x = duration_ms)) +
  geom_histogram(binwidth = 30000, color = "black", fill = "blue") +
  labs(title = "Histogram of Duration in Milliseconds", x = "Duration (ms)", y = "Frequency")

# Histogram for Danceability
ggplot(data = music_data, aes(x = danceability)) +
  geom_histogram(binwidth = 0.1, fill = "green", color = "black") +
  ggtitle("Histogram of Danceability") +
  xlab("Danceability") +
  ylab("Frequency")

# Histogram for Energy
ggplot(data = music_data, aes(x = energy)) +
  geom_histogram(binwidth = 0.1, fill = "red", color = "black") +
  ggtitle("Histogram of Energy") +
  xlab("Energy") +
  ylab("Frequency")

# Histogram for Instrumentalness
ggplot(data = music_data, aes(x = instrumentalness)) +
  geom_histogram(binwidth = 0.05, color = "black", fill = "green") +
  labs(title = "Histogram of Instrumentalness", x = "Instrumentalness", y = "Frequency")

# Histogram for Liveness
ggplot(data = music_data, aes(x = liveness)) +
  geom_histogram(binwidth = 0.05, color = "black", fill = "red") +
  labs(title = "Histogram of Liveness", x = "Liveness", y = "Frequency")

# Histogram for Loudness
ggplot(data = music_data, aes(x = loudness)) +
  geom_histogram(binwidth = 1, color = "black", fill = "purple") +
  labs(title = "Histogram of Loudness", x = "Loudness (dB)", y = "Frequency")

# Histogram for Speechiness
ggplot(data = music_data, aes(x = speechiness)) +
  geom_histogram(binwidth = 0.02, color = "black", fill = "orange") +
  labs(title = "Histogram of Speechiness", x = "Speechiness", y = "Frequency")

# Histogram for Tempo
ggplot(data = music_data, aes(x = tempo)) +
  geom_histogram(binwidth = 5, fill = "yellow", color = "black") + 
  ggtitle("Histogram of Tempo") +
  xlab("Tempo (BPM)") +
  ylab("Frequency")

# Histogram for Valence (Postiveness)
ggplot(data = music_data, aes(x = valence)) +
  geom_histogram(binwidth = 0.05, color = "black", fill = "yellow") +
  labs(title = "Histogram of Valence", x = "Valence", y = "Frequency")



```

Boxplots of the distributions of the Continuous Variables

```{r}
# BOXPLOT FOR REPONSE VARIABLE

# Boxplot for Popularity Scores
ggplot(data = music_data, aes(y = popularity)) +
  geom_boxplot(fill = "purple") +
  ggtitle("Boxplot of Popularity Scores") +
  ylab("Popularity Score")

```


```{r}
# BOXPLOTS FOR THE CONTINUOUS PREDICTOR VARIABLES IN THIS DATASET

# Boxplot for duration_ms
ggplot(music_data, aes(y = duration_ms)) +
  geom_boxplot(fill = "blue") +
  labs(title = "Boxplot of Duration in Milliseconds", y = "Duration (ms)", x = "") +
  theme_minimal()

# Boxplot for Danceability
ggplot(data = music_data, aes(y = danceability)) +
  geom_boxplot(fill = "green") +
  ggtitle("Boxplot of Danceability") +
  ylab("Danceability")

# Boxplot for Energy
ggplot(data = music_data, aes(y = energy)) +
  geom_boxplot(fill = "red") +
  ggtitle("Boxplot of Energy") +
  ylab("Energy")

# Boxplot for loudness
ggplot(music_data, aes(y = loudness)) +
  geom_boxplot(fill = "purple") +
  labs(title = "Boxplot of Loudness", y = "Loudness (dB)", x = "") +
  theme_minimal()

# Boxplot for speechiness
ggplot(music_data, aes(y = speechiness)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of Speechiness", y = "Speechiness", x = "") +
  theme_minimal()

# Boxplot for Acousticness
ggplot(data = music_data, aes(y = acousticness)) +
  geom_boxplot(fill = "blue") +
  ggtitle("Boxplot of Acousticness") +
  ylab("Acousticness")

# Boxplot for instrumentalness
ggplot(music_data, aes(y = instrumentalness)) +
  geom_boxplot(fill = "green") +
  labs(title = "Boxplot of Instrumentalness", y = "Instrumentalness", x = "") +
  theme_minimal()

# Boxplot for liveness
ggplot(music_data, aes(y = liveness)) +
  geom_boxplot(fill = "red") +
  labs(title = "Boxplot of Liveness", y = "Liveness", x = "") +
  theme_minimal()

# Boxplot for valence
ggplot(music_data, aes(y = valence)) +
  geom_boxplot(fill = "yellow") +
  labs(title = "Boxplot of Valence", y = "Valence", x = "") +
  theme_minimal()

# Boxplot for tempo
ggplot(music_data, aes(y = tempo)) +
  geom_boxplot(fill = "cyan") +
  labs(title = "Boxplot of Tempo", y = "Tempo (BPM)", x = "") +
  theme_minimal()





```

Scatterplots to explore the relationship between the Continuos predictor variables and the Reponse variable (popularity score)

```{r}
# Scatter plot for acousticness vs popularity
ggplot(music_data, aes(x = acousticness, y = popularity)) +
  geom_point() +
  labs(title = "Scatter plot of Acousticness vs Popularity", x = "Acousticness", y = "Popularity") +
  theme_minimal()

# Scatter plot for danceability vs popularity
ggplot(music_data, aes(x = danceability, y = popularity)) +
  geom_point() +
  labs(title = "Scatter plot of Danceability vs Popularity", x = "Danceability", y = "Popularity") +
  theme_minimal()

# Scatter plot for duration_ms vs popularity
ggplot(music_data, aes(x = duration_ms, y = popularity)) +
  geom_point() +
  labs(title = "Scatter plot of Duration (ms) vs Popularity", x = "Duration (ms)", y = "Popularity") +
  theme_minimal()

# Scatter plot for energy vs popularity
ggplot(music_data, aes(x = energy, y = popularity)) +
  geom_point() +
  labs(title = "Scatter plot of Energy vs Popularity", x = "Energy", y = "Popularity") +
  theme_minimal()

# Scatter plot for instrumentalness vs popularity
ggplot(music_data, aes(x = instrumentalness, y = popularity)) +
  geom_point() +
  labs(title = "Scatter plot of Instrumentalness vs Popularity", x = "Instrumentalness", y = "Popularity") +
  theme_minimal()

# Scatter plot for liveness vs popularity
ggplot(music_data, aes(x = liveness, y = popularity)) +
  geom_point() +
  labs(title = "Scatter plot of Liveness vs Popularity", x = "Liveness", y = "Popularity") +
  theme_minimal()

# Scatter plot for loudness vs popularity
ggplot(music_data, aes(x = loudness, y = popularity)) +
  geom_point() +
  labs(title = "Scatter plot of Loudness vs Popularity", x = "Loudness", y = "Popularity") +
  theme_minimal()

# Scatter plot for speechiness vs popularity
ggplot(music_data, aes(x = speechiness, y = popularity)) +
  geom_point() +
  labs(title = "Scatter plot of Speechiness vs Popularity", x = "Speechiness", y = "Popularity") +
  theme_minimal()

# Scatter plot for tempo vs popularity
ggplot(music_data, aes(x = tempo, y = popularity)) +
  geom_point() +
  labs(title = "Scatter plot of Tempo vs Popularity", x = "Tempo", y = "Popularity") +
  theme_minimal()

# Scatter plot for valence vs popularity
ggplot(music_data, aes(x = valence, y = popularity)) +
  geom_point() +
  labs(title = "Scatter plot of Valence vs Popularity", x = "Valence", y = "Popularity") +
  theme_minimal()

```
Correlation Analysis to check for multicollinearity between predictors

```{r}

continuous_vars <- music_data[, c("acousticness", "danceability", "duration_ms", "energy", "instrumentalness", "liveness", "loudness", "speechiness", "tempo", "valence", "popularity")]


correlation_matrix <- cor(continuous_vars, use = "complete.obs")  

print(correlation_matrix)



```

Using a variance inflation test to check for whether there's multicollinearity 

```{r}
## DONT RUN THIS

glm_model <- glm(popularity ~ acousticness + danceability + duration_ms +   energy + 
              instrumentalness + liveness + loudness + speechiness + tempo + 
              valence, data = music_data, family = poisson)

# Calculating VIF from the glm model
vif_values <- vif(glm_model, lp = glm_model$linear.predictors)

print(vif_values)
```





